Mergeable Types for High Availablity

### 2017-02-21 paper/model.tex
Model done.

  \begin{ocaml}
  module type DALI = sig
    type 'a 'b t
    val return : 'b -> 'a 'b t
    val bind : 'a 'b t -> ('b -> 'a 'c t) -> 'a 'c t
    val get_current_version: unit -> 'a 'a t
    val with_init_version_do: 'a -> 'a 'b t -> 'b
    val fork_version : 'a 'b t -> 'a unit t
    val sync ()_next_version: unit -> ?v:'a -> 'a 'a t
  end
  \end{ocaml}

@@

The \name programming model lets programmers define and compose
concurrent computations around versioned data structures.
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{'a 'b t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation async ()hronously
against the forked version.  The semantics of \C{sync ()\_next\_version}
(simply called \C{sync ()}) is somewhat involved: the function accepts a
\emph{proposal} for the next version of the state, where \C{next} is
defined w.r.t the current version, and returns (via a monad) the
actual next version, which becomes the current version for rest of the
computation.  The next version is created by merging the proposal with
a subset of concurrent versions that became available since the last
merge or fork. Thus, \C{sync ()} effectively lets a computation sync
() with a subset of concurrent computations and obtain their latest
updates.

@@

>>>>

  \begin{ocaml}
  module type DALI = sig
    type 'a 'b t
    val return : 'b -> 'a 'b t
    val bind : 'a 'b t -> ('b -> 'a 'c t) -> 'a 'c t
    val get_current_version: unit -> 'a 'a t
    val with_init_version_do: 'a -> 'a 'b t -> 'b
    val fork_version : 'a 'b t -> 'a unit t
    val sync_next_version: unit -> ?v:'a -> 'a 'a t
  end
  \end{ocaml}

@@

The \name programming model lets programmers define and compose
concurrent computations around versioned data structures.
Fig.~\ref{fig:dali-monad} shows the signature of the \name module that
implements the programming model along the lines of the well-known
\C{State} monad~\cite{wadler-monad}. The monad encapsulates a
versioned functional state (\C{'a}) and the type \C{'a 'b t}
represents a monadic computation that returns a \C{'b} result.
Functions \C{return} and \C{bind} have their usual monadic
interpretation. \C{get\_current\_version} is like the \C{State}
monad's \C{get}; it returns the versioned state encapsulated by the
monad. \C{with\_init\_version\_do} runs a monadic computation against
a given initial version and returns the result. \C{fork\_version}
returns a computation that forks a new concurrent version from the
current version, and runs the given monadic computation asynchronously
against the forked version.  The semantics of \C{sync\_next\_version}
(simply called \C{sync}) is somewhat involved: the function accepts
a \emph{proposal} for the next version of the
state, where \C{next} is defined w.r.t the current version, and
returns (via a monad) the actual next version, which becomes the
current version for rest of the computation.  The next version is
created by merging the proposal with a subset of concurrent versions
that became available since the last merge or fork. Thus, \C{sync}
effectively lets a computation sync with a subset of concurrent
computations and obtain their latest updates.

@@

The \drawsome example demonstrates the utility of mergeable data types
and the \name programming model in building concurrent applications
with conceptual sharing of state. This example, however, implicitly
makes certain assumptions about the model and the underlying system,
such as the existence of a single least common ancestor (LCA) for any
pair of versions, the ability to access a previous version on any
branch, and mergeability of any two concurrent versions. Enforcing
these guarantees in a fully decentralized distributed setting requires
addressing non-trivial theoretical challenges described in the next
section. Unlike the canvas, the process of writing merge functions for
data types with non-trivial invariants could be quite involved. The
later sections describe a principled approach to derive a merge
function for a data type given its invariants. Like \C{Canvas.merge},
merge functions often need to compare multiple versions of a data type
for structural equality, which, when performed naively, incurs cost
linear in the size of versions. The representation of a versioned data
type in \name is optimized for performing comparisons such that
structural equality can be determined in constant time. Such
engineering details, which contribute to the practicality of our
programming model are also discussed in subsequent sections.
